Online Appendices for ``GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations''

### 2025-07-02 writeUps/appendix.tex
finished?? editing

\documentclass[]{article}
\usepackage[letterpaper]{geometry}

\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
%\usepackage[notablist,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
\usepackage{threeparttable}
\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled

\usepackage[longnamesfirst,sort]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

\title{GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations}
%\runtitle{\textsc{geepers}}
%\maketitle

%\begin{aug}
 \author{
 Adam C. Sales\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu}, Kirk P. Vanacore, and Erin R. Ottmar%\\
 %\affil{Worcester Polytechnic Institute, Worcester, MA, USA}
}

% \maketitle

\maketitle

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that demonstrates the novel method's greater robustness compared to popular alternatives and illustrate the method through two real-data analyses.
\end{abstract}
% \begin{keywords}
% Causal Inference; Principal Stratification; Educational Technology
% \end{keywords}

\section{Introduction}

When students get stuck on math problems during homework or classwork, it may be helpful to offer them hints or other ``just-in-time'' support \cite[e.g.,][]{jit}, possibly helping them overcome specific misconceptions or other sources of confusion and allowing them to move on in the assignment.
The availability of problem-specific support for struggling students is one potential advantage of educational technology over traditional pencil-and-paper practice.
On the other hand, the ready availability of hints may discourage students from working through difficult problems independently, thereby hindering their learning.

The literature on hints in online learning is mixed \citep[see, e.g.,][]{aleven2016help,goldin2012learner,sales2021student}, particularly when it comes to requesting a ``bottom-out'' hint. In some online mathematics programs, sequences of hints, which a student can request one at a time, end with a bottom-out hint providing the answer to the problem. On the one hand, hints that include the correct answer are essentially worked examples, which can be beneficial to learning \citep[e.g.]{sweller1985use}. On the other hand, they allow students to ``game the system,'' by simply requesting enough hints to see the right answer and use the software without ever working to solve any problems \citep[e.g.][]{guo2008trying}.

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate.
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints.
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that the availability of bottom-out hints played a role in the condition's effectiveness.
Alternatively, if students who avoided bottom-out hints benefited more, it may be because the availability of bottom-out hints impeded the condition's effectiveness.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)
Indeed, if the use of bottom-out hints is considered part of the treatment, then those students who rarely used them were partial compliers---did the treatment still benefit them?

Estimating an effect for bottom-outers, or other partial compliers, is a tricky proposition.
Causal inference in statistics typically depends on comparisons of outcomes between similar groups of subjects. However, there is no readily identifiable comparison group for bottom-outers, who are only represented in one of the four treatment conditions, and who presumably struggled more frequently, on average, than their non-bottom-outer peers.
We will formulate a solution using principal stratification \citep{frangakis}, a framework for defining and estimating effects for post-treatment subgroups, that is, subgroups that may themselves be a function of treatment assignment. It relies on the notion of \emph{potential} subgroups called principal strata: which subgroup would a subject be in if they were (perhaps counterfactually) assigned to one treatment condition or another.
For example, the appropriate comparison group for bottom-outers is the set of students, assigned to the other treatment arms, who \emph{would have} requested bottom-out hints if only given that option.

Since principal strata membership is based on subjects' responses to both factual and counterfactual treatment assignments, it is unobserved---we don't observe which students in the conditions without bottom-out hints would have bottomed out if given the opportunity.
Nonetheless, researchers have devised estimators of principal effects---average effects within principal strata---to address a broad range of applied statistical problems.

\label{background}The conceptual basis for principal stratification---though not the terminology---emerged from the study of randomized experiments with noncompliance. Specifically, \citet{air} showed that the econometric instrumental variables estimand is the average effect of the treatment on the principal stratum of ``compliers''---those subjects who would receive the treatment if and only if randomized to the treatment condition. Moreover, the average effect for compliers is nonparametrically identified if a random treatment assignment only has an effect for compliers---``exclusion restriction''---and if treatment assignment never \emph{prevents} any subjects from receiving treatment (``monotonicity'').
In a follow-up analysis, \citet{imbens1997bayesian} relaxed the exclusion restriction by way of parametric Bayesian finite mixture modeling, which remains a popular approach.

\citet{frangakis} introduced the terminology of principal stratification and extended the framework well beyond noncompliance, inspiring a broad literature introducing (and debating) novel applications and estimators.
For instance, \citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program.
Other methodologists have developed estimators for evaluating surrogate outcomes \citep{li2010bayesian}, investigating causal mechanisms \citep{lidsayPage}, addressing sample attrition \citep{zhangRubin, ding2011}, and other related problems.
Since interest is often in comparing effects between principal strata---a goal that runs counter to the exclusion restriction---the earliest principal stratification estimators built on \citet{imbens1997bayesian}'s parametric mixture models \citep[e.g.][]{Barnard01062003,mealli2004analyzing}, and that approach remains popular \citep{lidsayPage,fellerEtAl2016}.
However, these models are often difficult to specify and can be either hard or impossible to test.
Even when they are well-specified, they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

These concerns have prompted the development of alternative approaches, including bounding \citep{bounding} and randomization inference \citep{nolen2011randomization}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute an unbiased estimate of principal effects.
Principal ignorability is a direct analogue of the strong ignorability assumption in causal observational studies \citep[e.g.][]{rosenbaum1983central}---that all covariates that predict both principal stratum membership and outcomes have been accounted for---in other words, the precise type of assumption that randomized trials are designed to avoid.
A separate set of results in \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally, estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

After the initial submission of this manuscript, we became aware of \citet{richardson2023estimating}, which suggests a similar, though not identical, estimator to ours, using G-estimation of a structural nested mean model \citep[e.g.][]{wallace2017r}.
Our development of \textsc{geepers} builds on \citet{richardson2023estimating} by emphasizing connections to both parametric mixture modeling and principal score analysis, and comparing \textsc{geepers} to those methods in a simulation study.
Moreover, \textsc{geepers}, which, in practice, is a combination of logistic regression and OLS, is more accessible to applied researchers while yielding very similar results.

This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method, as well as the trade-offs between them.

The following section gives a formal introduction to principal stratification and briefly describes two common estimation techniques, parametric mixture modeling and principal score weighting.
Next, Section \ref{sec:geepers} develops \textsc{geepers} estimation and gives conditions for its consistency.
Section \ref{sec:simulation} describes the simulation study. % examining the operating characteristics of \textsc{geepers} estimates and comparing them to estimates from parametric mixture modeling and principal score weighting.
%Under ideal circumstances, estimates from parametric mixture models and weighting estimators are somewhat more efficient than \textsc{geepers}, but when assumptions are violated \textsc{geepers} easily out-performs the other two methods.
Section \ref{sec:fh2t} illustrates \textsc{geepers} in two applied examples: first, we re-analyze the applied example from \citet{richardson2023estimating}, a randomized trial of a periodontal treatment for pregnant women in which there was significant one-way noncompliance. Second, we return to the question of bottom-out hints: does the effect of being randomized to a condition that includes bottom-out hints depend on how often they are requested?  Section \ref{sec:discussion} concludes.

\section{Background}
In a randomized experiment with two conditions, let $Z_i=1$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken after treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff0=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

\subsection{Estimating Principal Effects}
In estimating principal effects for a binary $S$, it will be useful to define four conditional means. For $s\in\{0,1\}$, let
\begin{equation}\label{eq:mus}
\mucs=\EE[Y_C|\st=s] \mbox{ and }\muts=\EE[Y_T|\st=s]%\mbox{ for }s=0,1
  % \begin{split}
  %   \muc0&=\EE[Y_C|\st=0]\\
  %   \muc1&=\EE[Y_C|\st=1]\\
  %   \mut0&=\EE[Y_T|\st=0]\\
  %   \mut1&=\EE[Y_T|\st=1]
  % \end{split}
\end{equation}
Then, since
\begin{equation*}
  \effs=\muts-\mucs
  % \begin{split}
  %   \eff0&=\mut0-\muc0\\
  %   \eff1&=\mut1-\muc1
  % \end{split}
\end{equation*}
estimating principal effects requires estimating $\muc0$, $\mut0$, $\muc1$, and $\mut1$.

We will focus on estimation in completely randomized experiments; that is, we will assume
\begin{ass}[Randomization]\label{ass:rand}
\begin{equation*}%\label{eq:randomization}
  \yci,\yti,\sti, \bm{x}_i \independent Z_i
\end{equation*}
\end{ass}
where $\bm{x}_i$ is a vector of pre-treatment covariates, and $\independent$ denotes independence.

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are nonparametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
% When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
% In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$% that do not assume the exclusion restriction
---normal mixture modeling and weighting---with some comments in-between on the role for covariates.

\subsubsection{Parametric Mixture Modeling}
%The normal mixture modeling approach to estimating $\muc0$ and $\muc1$ \citep{imbens1997bayesian} assumes that, conditional on $\st$, $Y_C$ is normally distributed.
Say an analyst assumes that, within principal strata, $Y_C$ is drawn from a parametric model, i.e., $f_{Y_C}(y|\st=s)=g^s(y)$. For instance, \citep{imbens1997bayesian} assumes $Y_C$ is normally distributed within strata, so $g^s(y)=\phi\left(\frac{y-\mucs}{\sigma_C^s}\right)$, where $\phi(\cdot)$ is the standard normal density function, and $\sigma_C^0>0$ and $\sigma_C^1>0$ are standard deviations. 
Then the probability density of $Y_C$ may be written as
\begin{equation}\label{eq:mixtureModel}
 % f_{Y_C}(y)=Pr(\st=0)\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+Pr(\st=1)\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
 f_{Y_C}(y)=Pr(\st=0)g^0(y)+Pr(\st=1)g^1(y).
\end{equation}
The probabilities $Pr(\st=0)$ and $Pr(\st=1)$ may be estimated first using data from the treatment group, since Assumption \ref{ass:rand} implies that $Pr(\st=1|Z=1)=Pr(\st=1)$.
Given these estimates and \eqref{eq:mixtureModel}, parameters $\muc0$ and $\muc1$ may be estimated using maximum likelihood, method of moments, or Bayesian techniques.

This approach has several problems.
First, the assumption of a conditional parametric model is untestable and limiting.
Unfortunately, even when it holds, mixture model estimates can be unreliable and, in some cases, severely biased \citep{griffin2008application,feller2017principal}.

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi),
\end{equation}
%where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
the probability of a subject with covariates $\bxsi$ being in the $\sti=1$ principal stratum.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ as $\bm{\hat{\alpha}}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\hppi=\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=0$.

Going forward, we will occasionally suppress dependence on $\bm{\alpha}$, and write $\ppip{\bm{\alpha}}=\ppi$.

If $\bm{x}$ is informative about $\st$, principal scores will vary between subjects. We will assume this is the case: %the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
%In some estimators we will discuss below, two distinct values suffice.
For a broader discussion, see \citet{ding2011,jiangDing2021}.

Principal scores can potentially improve inference in a finite mixture model by modifying \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y|\bm{x})=(1-\ppip{\bm{\alpha}})%\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)
g^0(y|\bm{x})+\ppip{\bm{\alpha}}%\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
g^1(y|\bm{x})
\end{equation}
where $g^s(y|\bm{x})$, $s=0,1$, are conditional densities which must be specified. Most commonly, researchers assume a linear normal model, e.g., $g^s(y|\bm{x})=\phi\left(\frac{y-\mucs-\bm{\gamma}'\bm{x}}{\sigma_C^s}\right)$, where the columns of $\bm{x}$ are centered so that $\mathbb{E}[\bm{x}]=\bm{0}$. 
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo,dingLu}:
%\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
%\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies. % \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$;
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

Under principal ignorability, $\muc0$ and $\muc1$ can be estimated via weighting:
\begin{equation}\label{eq:psw}
  \hat{\muc0}_w=\frac{\sum_{i: Z_i=0} Y_i(1-\ppi)}{\sum_{i:Z_i=0}1-\ppi}\text{ and } \hat{\muc1}_w=\frac{\sum_{i: Z_i=0} Y_i(\ppi)}{\sum_{i:Z_i=0}\ppi}
\end{equation}

Principal score weighting (\textsc{psw}) does not require any distributional assumptions and (after estimating principal scores) is very easy to implement.
On the other hand, the principal ignorability assumption is strong and restrictive.

\citet{feller2017principal} recommends the case-resampling bootstrap to estimate the sampling variances of $\hat{\mu}_{Cw}^0$, $\hat{\mu}_{Cw}^1$, and their associated principal effect estimators.

\subsection{M-Estimation}\label{sec:mest}
M-estimation (also generalized method of moments or generalized estimating equations) is a general approach to deriving consistent statistical estimators and their asymptotic distributions.
In the following section, we will describe \textsc{geepers}, a novel principal effect estimator, and derive its properties using the principles of M-estimation.
Here, we provide a brief overview of M-estimation, primarily drawn from the excellent treatment in \citet{stefanskiBoos}.

Let $\bm{\mathcal{O}}_i$ be a vector of observed data for subject $i=1,\dots,n$, with $\bm{\mathcal{O}_1},\dots,\bm{\mathcal{O}}_n\overset{\mathrm{iid}}{\sim} P$ for some distribution $P$, %\bm{\mathcal{O}}_i\independent \bm{\mathcal{O}}_{i'}$ if $i\ne i'$, 
let $\bm{\theta}$ be a vector of parameters, 
and let $\bm{\psi}(\bm{\mathcal{O}},\bm{\theta})$ be a vector field such that $\mathbb{E}_P[\bm{\psi}(\bm{\mathcal{O}},\bm{\theta}_0)]=\bm{0}$ for the true parameter value $\bm{\theta}_0$. 
Then, an M-estimator $\bm{\hat{\theta}}$ satisfies ``estimating equations'' $\sum_i \psi_j\left(\bm{\mathcal{O}}_i,\bm{\hat{\theta}}\right)=0$. 
Some common examples are maximum likelihood estimators, where $\bm{\psi}(\bm{\mathcal{O}},\bm{\theta})$ is the score function %, with $\psi_j(\bm{\mathcal{O}},\bm{\theta})=\partial log f(\bm{\mathcal{O}},\bm{\theta})/\partial \theta_j$ and $f(\cdot,\bm{theta})$ is the density of $\bm{\mathcal{O}}$, 
or method of moments estimators, in which, if data $\mathcal{O}$ are unidimensional, $\psi(\mathcal{O},\bm{\theta})=\mathcal{O}-\EE_\theta\left[\mathcal{O}\right]$.

Under suitable regularity conditions, an M-estimator $\bm{\hat{\theta}}$ is asymptotically normal, and its sampling variance-covariance matrix may be estimated using the formula \citep[][ch. 7]{boosStefanskiBook}:
\begin{equation}\label{eq:sandwich}
  \widehat{var}(\bm{\hat{\theta}})=ABA'
\end{equation}
%\citep[][ch. 7]{boosStefanskiBook}.
with matrices
\begin{equation*}
    A=\left\{\frac{1}{n}\displaystyle\sum_i\left. \frac{\partial \bm{\psi}(\bm{\mathcal{O}_i},\bm{\theta})'}{\partial \bm{\theta}'}\right|_{\bm{\theta}=\bm{\hat{\theta}}}\right\}^{-1}\mbox{ and }B=\frac{1}{n}\sum_i \bm{\psi}(\bm{\mathcal{O}_i},\bm{\hat\theta})\bm{\psi}(\bm{\mathcal{O}_i},\bm{\hat\theta})'.
\end{equation*}
Because of its form, \eqref{eq:sandwich} is sometimes called a ``sandwich'' formula, where $A$ is the ``bread'' and $B$ is the ``meat'' \citep[e.g.][]{sandwich}. 
%To define the matrices $A$ and $B$, let the vector $\bm{\psi}_i=\left\{\psi_j(\bm{\mathcal{O}}_i,\bm{\hat{\theta}})\right\}_{j=1}^J$. Then,
% \begin{equation*}
%   A^{-1}=\frac{1}{n}\sum_i\left. \frac{\partial \bm{\psi}_i'}{\partial \bm{\theta}'}\right|_{\bm{\theta}=\bm{\hat{\theta}}}
% \end{equation*}
% and
% \begin{equation*}%\label{eq:Bmat}
%   B=\frac{1}{n}\sum_i \bm{\psi}_i\bm{\psi}_i'
% \end{equation*}
In general, the conditions for asymptotic normality may be difficult to prove; fortunately, they are met for OLS and common generalized linear models such as logistic regression, so long as the dimension of $\bm{\theta}$ does not increase with $n$.

Multiple M-estimators can be combined by ``stacking'' their estimating equations.
In particular, say an M-estimator with estimating equations $\bm{\psi}^2(\bm{\mathcal{O}},\bm{\hat\theta^2})=\bm{0}$ takes, as input, the results from a previous M-estimator, with estimating equations $\bm{\psi}^1(\bm{\mathcal{O}},\bm{\hat\theta^1})=\bm{0}$.
The resulting estimator is the solution to the combined estimating equations
\begin{equation}\label{eq:stacked}
  \psi(\bm{\mathcal{O}},[\bm{\hat\theta^1},\bm{\hat{\theta}^2}])=[\psi^1(\bm{\mathcal{O}},\bm{\hat\theta^1}),\psi^2(\bm{\mathcal{O}},\bm{\hat\theta^2})]=\bm{0},
\end{equation}
%where the superscript $s$ stands for ``stacked.''
A variance-covariance matrix estimated using the sandwich formula based on $\psi$ accounts for uncertainty from both steps of the estimation procedure.

\section{GEEPERS }\label{sec:geepers}

The approach we introduce here incorporates principal scores into an M-Estimator for the mixture model \eqref{eq:mixtureModel}.
Like principal score weighting, our M-estimator does not require distributional assumptions but instead requires a conditional independence assumption.
We will begin by describing a stronger-than-necessary conditional independence assumption to build intuition; in \S~\ref{sec:regression} we will present a considerably weaker alternative.

\subsection{Building Intuition: Covariate Ignorability}\label{sec:ci}

We begin by introducing an assumption that we term ``Covariate Independence,'' or CI. It is a slightly weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Independence]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e., $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances, researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS. % assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

The argument stems from a set of expressions for expectations, summarized in the following lemma:
\begin{lemma}\label{lemma:expectation}
Let \begin{equation}\label{eq:estEq0}
\tilde{\Psi}_i=\begin{pmatrix}
    (1-Z_i)\left[Y_i-\muc0-\ppi(\muc1-\muc0)\right]\\
    (1-Z_i)\left[\ppi Y_i-\ppi\muc0-\ppi^2(\muc1-\muc0)\right]\\
    Z_i\left[Y_i-\mut0-\sti (\mut1-\mut0)\right]\\
    Z_i\left[\sti Y_i -\sti\mut0-\sti^2(\mut1-\mut0)\right]
  \end{pmatrix}
\end{equation}
Then under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, $\EE[\Psi_i]=\bm{0}$.
\end{lemma}

Lemma \ref{lemma:expectation} implies a set of estimating equations $\sum_{i=1}^n\tilde{\Psi}_i=\bm{0}$.

Now, the estimating equations $\sum_i\tilde{\Psi}_i=\bm{0}$ are algebraically equivalent to the estimating equations for a particular OLS fit, after some transformations, the most important of which follows.
Let
\begin{equation}\label{eq:ri}
\ri=Z_iS_i+(1-Z_i)\ppi=\begin{cases}
\ppi &Z_i=0\\
\sti &Z_i=1
\end{cases}
\end{equation}
In general, $(1-Z_i)\ppi=(1-Z_i)\ri$ and $Z_i\sti=Z_i\ri$, allowing $\ri$ to take the place of $\ppi$ and $S_i$ throughout $\tilde{\Psi}_i$.
Denote estimating equations based on $\ri$ as $\sum_i \psi_i=\bm{0}$.

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  %\begin{split}
    \heff0_{CI}\equiv \hat{\beta}_2\mbox{ and }
    \heff1_{CI}\equiv \hat{\beta}_3+\hat{\beta}_2
%  \end{split}
\end{equation}
are M-estimators, with $\heff0_{CI}\rightarrow_p\eff0$ and $\heff1_{CI}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

Thus, given principal scores, Assumption \ref{ass:ci} enables an analyst to estimate principal effects with a simple regression.
Proposition \ref{prop:reg1} builds on Theorem 2 of \citet{jiangDing2021} which establishes identification of $\eff0$ and $\eff1$ by representing the principal effect estimators as a familiar regression using estimated principal scores.

% If principal scores were known exactly, the M-estimation structure of $\heff0$ and $\heff1$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
% This will rarely be the case.
%However,
If the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_i Z_i\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked as in \eqref{eq:stacked}.
Sandwich standard error estimates of the form \eqref{eq:sandwich} based on $\left[Z_i\omega_i,\psi_i\right]$ will incorporate uncertainty from both principal score modeling and principal effect estimation conditional on principal scores.

\subsection{Relaxing Covariate Independence with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).

The assumption we'll state here is stronger than necessary, but leads to an attractively simple method; a weaker version of the assumption and corresponding method, along with the proof, can be found in an appendix.

\begin{assp}{\ref*{ass:ci}$'$}[Residualized Covariate Independence]\label{ass:rci}
There exists $\bxy$, a (known) multivariate transformation of covariates $\bx$, and an (unknown) vector of coefficients $\bm{\gamma}$ such that
\begin{equation}\label{eq:ass:rci}
\begin{split}
\EE[\yc-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yc-\bm{\gamma}'\bxy|\st]\\
\EE[\yt-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yt-\bm{\gamma}'\bxy|\st]
\end{split}
\end{equation}
\end{assp}
That is, CI may not hold for potential outcomes themselves, but an analogous assumption holds for \emph{residualized} potential outcomes---i.e., $\yc$ and $\yt$ after subtracting out a linear function of (possibly transformed) covariates.
By forcing $\bm{\gamma}$ to be the same for both treatment groups, and (implicitly) for both principal strata, Assumption \ref{ass:rci} assumes that there are no interactions between columns of $\bxy$ and either condition or principal stratum.
This is akin to the ``Additivity of Treatment Assignment Effect'' assumption in \citet{jo2002}.
Proposition \ref{prop:interactions} in the appendix relaxes these requirements, but some preliminary simulations suggest that standard errors from estimates allowing for these interactions will be prohibitively large.
The simulation results in the following section address cases in which interactions are present in the data-generating model but not in the data analysis.

In any event, Assumption \ref{ass:rci} allows principal effects to be estimated with an OLS model including $\bxy$:
\begin{prop}{\textsc{geepers}}\label{prop:reg2}

>>>>

\documentclass[]{article}
\usepackage[letterpaper]{geometry}
\usepackage{xr}
\externaldocument{writeUps/paper}
\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage[caption=false]{subfig}
\usepackage[doublespacing]{setspace}

\usepackage[longnamesfirst,sort]{natbib}
\bibpunct[, ]{(}{)}{;}{a}{,}{,}%
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

\title{Online Appendices for ``GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations''}
%\runtitle{\textsc{geepers}}
%\maketitle
